Variana Reloaded

### 2022-03-07 cv/cv.tex
Updated cv

@@

\noindent Hands-on scientific team leader, graduated from \'Ecole Centrale Paris, PhD in engineering science, with a specialist expertise in computer vision, data science and machine learning. Has operated in the fields of academic research, healthcare, electro-domestic and entertainment industry. Strong experience in the development of both open source and commercial software. {\em Keywords:} applied mathematics, artificial intelligence, image processing, computer vision, medical image analysis, machine learning, deep learning, statistics, Python,~C.

@@

\noindent \textbf{DATA SCIENCE:} In-depth knowledge of statistics, computer vision, medical image analysis, traditional machine learning and deep learning.

\ \\
\noindent \textbf{SCIENTIFIC COMPUTING:} Over ten years experience in Python programming using the NumPy/SciPy ecosystem and interfacing C~language with Python.

@@

\bighead{HEAD OF COMPUTER VISION \& ARTIFICIAL INTELLIGENCE}{Didimo, Porto, Portugal}{2019--present}
\nopagebreak
\begin{itemize}
\item Developing computer vision and machine learning algorithms for reconstruction of high-fidelity digital humans from photos.
\end{itemize}
\ \\

\bighead{SENIOR COMPUTER VISION SCIENTIST}{CoVii, Ar\c celik/Beko group, Porto, Portugal}{2017--2019}
\nopagebreak
\begin{itemize}
\item Developed embedded object recognition and tracking algorithms for smart domestic appliances using traditional computer vision and deep learning (VUXHub and Artisan applications demonstrated at IFA Berlin, 2017--18).
\end{itemize}

\ \\
\bighead{LEAD CLINICAL RESEARCH -- ADVANCED CLINICAL IMAGING TECHNOLOGY}{Siemens Healthineers / Lausanne University Hospital (CHUV), Switzerland}{2011--2017}
\nopagebreak
\begin{itemize}
\item Led algorithmic development of the {\em MorphoBox} Siemens prototype for brain morphometry using anatomical MRI to help radiological reading for patients with suspected neurodegeneration.
\item Contributed a number of clinical validation studies of brain morphometry in collaboration with neurologists and radiologists, via statistical analysis and data mining techniques, in Alzheimer's disease and other dementia, mild cognitive impairment, multiple sclerosis, migraine, anorexia, \ldots 
\item Co-supervised three PhD students and two MSc students.
\item Main organizer of annual Siemens/CHUV {\em brain imaging} workshops in Lausanne, Switzerland, 2012--2016.
\end{itemize}

\ \\
\bighead{PERMANENT RESEARCHER -- NEUROIMAGING INSTITUTE}{French Atomic Commission (CEA), Paris, France}{2002--2011}
\nopagebreak
\begin{itemize}
\item Developed advanced algorithms for image processing and statistical analysis of functional, anatomical and diffusion-weighted MRI data (population analysis, spatio-temporal image registration, real-time imaging).
\item French ANR-funded project management: PI (Karametria project, statistical analysis of brain structures, budget: 620 K\euro, 2009--2011), team leader (NIBB project, studying language in infants via functional neuroimaging, 2006--2009).
\item Active contributor to the NiPy software library (Neuroimaging in Python, \url{www.nipy.org}) from 2006.
\item Main supervisor of two PhD students and four MSc students.
\item Academic Guest at Computer Vision Laboratory, Swiss Federal Institute of Technology Zurich (ETHZ), Switzerland, 2009--2011.
\end{itemize}

@@

@@

\heading{Education}
\nopagebreak
\head{PhD, Engineering Science (1997--2001)}{French National Research Institute (INRIA) \& University of Nice-Sophia Antipolis, France}
\begin{itemize}
\item Development of multimodal image registration algorithms, with applications in radiotherapy, image-guided surgery and neuroscience. With highest honor.
\end{itemize}

@@

\ \\
\head{Engineer Degree (equivalent MSc) (1993--1996)}{Ecole Centrale Paris (now CentraleSupelec), France}
\nopagebreak
\begin{itemize}
\item Third year specialization in Applied Mathematics.
\end{itemize}

@@

\heading{Additional information}
\nopagebreak
\noindent 
\begin{minipage}{.15\textwidth}
\textbf{Languages:}
\end{minipage}
\begin{minipage}{.85\textwidth}
Native French, fluent English, basic Italian \& Portuguese. %German.
\end{minipage}

\ \\
\noindent 
\begin{minipage}{.15\textwidth}
\textbf{Computer skills:}
\ \\ 
\ \\
\ \\ 
\ \\
\end{minipage}
\begin{minipage}{.85\textwidth}
{\tiny$\bullet$} Programming languages: C, Python, Matlab, R. \\
{\tiny$\bullet$} Python packages: numpy, scipy, pylab, skimage, opencv, sklearn, pytorch, tensorflow, keras.\\
{\tiny$\bullet$} C/Python integration via cython.\\
{\tiny$\bullet$} Version control software: git, svn, perforce. \\
{\tiny$\bullet$} Reporting: \LaTeX, MS~Office, OpenOffice, LibreOffice, Google~Docs. \\
\end{minipage}

\ \\
\noindent 
\begin{minipage}{.15\textwidth}
\textbf{Publications:}
\ \\ 
\ \\
\ \\
\end{minipage}
\begin{minipage}{.85\textwidth}
{\tiny$\bullet$} 115 scholarly publications. Full list: \url{http://orcid.org/0000-0002-4821-6893}. \\
{\tiny$\bullet$} Google scholar statistics: 6602 citations, h-index: 33, i10-index: 66 (July 2021).\\
{\tiny$\bullet$} 5 published international patents. 2 ongoing applications.\\
\end{minipage}

%\ \\
%\noindent 
%\begin{minipage}{.15\textwidth}
%\textbf{ Reviewing:}
%\ \\ 
%\ \\
%\end{minipage}
%\begin{minipage}{.85\textwidth}
%IEEE Transactions on Medical Imaging, Medical Image Analysis, NeuroImage, Frontiers in
% Neuroscience, IEEE Transactions on Image Processing, Signal Image and Video Processing, MICCAI conference, \ldots
%\end{minipage}

\ \\
\noindent 
\begin{minipage}{.15\textwidth}
\textbf{Academic work:}
\ \\ 
\ \\
\ \\
\ \\
\ \\
\end{minipage}
\begin{minipage}{.85\textwidth}
  {\tiny$\bullet$} Post-graduate teaching in computer vision and medical image analysis: about 100~hours (Ecole Centrale Paris, EPFL Lausanne, University of Nice, University of Lausanne, INSERM/CNRS continuing education, \ldots).\ \\
  {\tiny$\bullet$} Long-time scientific journal reviewing experience (IEEE Trans.~Medical Imaging, Medical Image Analysis, NeuroImage, IEEE Trans.~PAMI, IEEE Trans.~Signal Processing, Frontiers in Neuroscience, \ldots).\ \\  
\end{minipage}

\ \\
\noindent 
\begin{minipage}{.15\textwidth}
\textbf{References:}
\end{minipage}
\begin{minipage}{.85\textwidth}
Available on request.
\end{minipage}

>>>>

\newcommand{\tabitem}{~~\llap{\textbullet}~~}

@@

\noindent Hands-on scientific team leader, graduated from \'Ecole Centrale Paris, PhD in engineering science, with a specialist expertise in computer vision and machine learning. Has operated in academic research for 15 years before moving to R\&D in industry (healthcare, electro-domestic, entertainment). Strong experience in the development of both open source and commercial software in Python, C, C++. {\em Keywords:} applied mathematics, statistics, artificial intelligence, machine learning, computer vision, medical image analysis, clinical studies, computer graphics.

@@

\noindent \textbf{STATISTICAL MODELING:} Strong experience developing model-based algorithms for statistical inference, with applications in machine learning, computer vision, computer graphics, \ldots

\ \\
\noindent \textbf{SCIENTIFIC COMPUTING:} 15 years experience in Python programming within the NumPy/SciPy ecosystem and interfacing C~language with Python.

@@

\bighead{HEAD OF COMPUTER VISION \& ARTIFICIAL INTELLIGENCE}{Didimo, Porto, Portugal}{2019--present}
\nopagebreak
\begin{itemize}
\item Leading development of algorithms for automated generation of high-fidelity digital humans from photos. {\em Keywords:} facial reconstruction, ``deep face'', animation retargetting, image enhancement.
\end{itemize}
\ \\

\bighead{SENIOR COMPUTER VISION SCIENTIST}{CoVii, Ar\c celik/Beko group, Porto, Portugal}{2017--2019}
\nopagebreak
\begin{itemize}
\item Developed embedded algorithms for smart domestic appliances (VUXHub and Artisan intelligent oven prototypes demonstrated at IFA Berlin, 2017--18). {\em Keywords:} image classification, object recognition \& tracking, deep learning. 
\end{itemize}
\ \\

\bighead{LEAD CLINICAL RESEARCH -- ADVANCED CLINICAL IMAGING TECHNOLOGY}{Siemens Healthineers / Lausanne University Hospital (CHUV), Switzerland}{2011--2017}
\nopagebreak
\begin{itemize}
\item Led algorithmic development of a brain morphometry tool (released as part of the {\em AI-Rad Companion} Siemens solutions) to help radiological reading for patients with suspected neurodegeneration. {\em Keywords:} magnetic resonance imaging (MRI), brain image segmentation, machine learning for disease classification. 
\item Main organizer of annual Siemens/CHUV {\em brain imaging} workshops in Lausanne, Switzerland, 2012--2016.
\end{itemize}

\ \\
\bighead{PERMANENT RESEARCHER -- NEUROIMAGING INSTITUTE}{French Atomic Commission (CEA), Paris, France}{2002--2011}
\nopagebreak
\begin{itemize}
\item Developed advanced algorithms for image processing and statistical analysis of brain imaging data. {\em Keywords:} functional, anatomical and diffusion-weighted MRI, image-based population analysis, spatio-temporal image registration, real-time imaging.
\item Funded project management: principal investigator ({\em Karametria} -- statistical analysis of brain structures, French National Research Agency project, budget: 620 K\euro, 2009--2011); team leader ({\em NIBB} -- studying language in infants via functional neuroimaging, French National Research Agency project, 2006--2009).
\item Active contributor to the NiPy software library (Neuroimaging in Python, \url{www.nipy.org}) from 2006.
\item Academic Guest at Computer Vision Laboratory, Swiss Federal Institute of Technology Zurich (ETHZ), Switzerland, 2009--2011.
\end{itemize}

@@

\ \\
\bighead{PhD CANDIDATE -- EPIDAURE PROJECT}{French National Research Institute (INRIA), Sophia Antipolis, France}{1997--2001}
\nopagebreak
\begin{itemize}
\item Development of multimodal image registration algorithms, with applications in radiotherapy, image-guided surgery and neuroscience. 
\end{itemize}

@@

\heading{Education}
\nopagebreak
\head{PhD, Engineering Science (1997--2001)}{University of Nice-Sophia Antipolis, France}
\begin{itemize}
\item With highest honor.
\end{itemize}

@@

\ \\
\head{Engineer Degree (equivalent MSc) (1993--1996)}{Ecole Centrale Paris, France}
\nopagebreak
\begin{itemize}
\item Third year specialization in Applied Mathematics.
\end{itemize}

@@

\heading{Additional information}
\nopagebreak
\noindent

\begin{tabular}{p{0.2\linewidth}p{0.8\linewidth}}
  \textbf{Languages:} & Native French, fluent English, basic Italian \& Portuguese.\\
  \hline
  \textbf{IT skills:} &
  \tabitem Programming languages: Python, C, C++, Matlab, R.\\
  & \tabitem Long-term experience with scientific Python pacakges (numpy, scipy, pylab).\\
  & \tabitem CV/AI Python packages: skimage, PIL, opencv, sklearn, pytorch, tensorflow, keras.\\
  & \tabitem  C/Python integration via Cython.\\
  & \tabitem Version control software: git, svn, perforce.\\
  & \tabitem Agile workflow (mainly Scrum) using JIRA \& Confluence.\\
  & \tabitem Reporting: \LaTeX, MS~Office, OpenOffice, LibreOffice, Google~Docs.\\
  \hline
  \textbf{Publications:} &
  \tabitem ORCID: \url{http://orcid.org/0000-0002-4821-6893}. \\
  & \tabitem 115 scholarly publications.\\
  & \tabitem Google scholar statistics: 6862 citations, h-index: 33, i10-index: 68 (March 2022).\\
  & \tabitem 5 published international patents. 2 ongoing applications.\\
  \hline
  \textbf{Academic work:} & 
  \tabitem Post-graduate teaching in computer vision and medical image analysis: about 100~hours (Ecole Centrale Paris, EPFL Lausanne, University of Nice, University of Lausanne, INSERM/CNRS continuing education, \ldots).\\
  & \tabitem Long-time scientific journal reviewing experience (IEEE Trans.~Medical Imaging, Medical Image Analysis, NeuroImage, IEEE Trans.~PAMI, IEEE Trans.~Signal Processing, Frontiers in Neuroscience, \ldots).\\
  & \tabitem Supervision: two PhD students and four MSc students as main supervisor, three PhD students and two MSc students as co-supervisor.\\
  \hline
  \textbf{References:} & Available on request.
\end{tabular}

### 2022-03-07 papers/variana/notes.tex
Updated cv

@@

The other option is to approximate the KL gradient altogether:
$$
\nabla D \approx {\cal I}[\pi(g_n - f)\phi].
$$
This is where the variational sampling story may start. Option~2, which corresponds to our initial variational sampling idea, is arguably better because it is exact whenever $g_n=f$, meaning that the Newton iteration will converge deterministically whenever $f$ is in the approximating family. Hence, for a near Gaussian target, we expect less numerical instabilities with Option~2. Note that, at this level of description, there is no sampling kernel involved -- this is a specific detail of numerical integration.

\section{Naive VS}

What the variance inequalities essentially prescribe is that the target factor~$f$ be a fast decreasing bounded function (it should not have heavier tails than polynomial in the Gaussian case). In practice, this constraint is easily fulfilled, for instance if~$f$ is a posterior distribution arising from an exponential prior and a bounded likelihood function. Other compliant factors can be derived from this case, as we will see. However, importance weighting functions resulting from distribution ratios are ruled out.

Consider a Bayesian posterior distribution of the form $p=\pi L$, where $\pi$ is a prior and $L$ a likelihood function. A natural approach would be to apply VS with $\pi$ as the sampling distribution and $f=p/\pi=L$ as the factor to be approximated. This means that we are trying to approximate the likelihood function by a Gaussian function in the context of the prior. Makes sense. This is the canonical approach exposed in my previous publications on VS.

>>>>

\section{VS vs IS}

The classical Monte Carlo approach to approximate an integral $\int p$ is to sample from some kernel $\pi$ and compute the empirical mean of importance weights $p(x_i)/\pi(x_i)$. This is known as importance sampling, but it's just the old Monte Carlo integration idea if we start from the form $p=\pi f$ and realize that the importance weighting function is then nothing but the target factor. The IS estimator is just a sample mean:
$$
\hat{I}(f) = \frac{1}{n}\sum_i f(x_i),
\qquad x_i \stackrel{iid}{\sim} \pi(x),
$$
which has the same mean and $1/n$ times the variance of the random variable $f(X)$, for $X\sim\pi(x)$. Beside the obvious fact that:
$$
\mathbb{E}[f(X)] = \int\pi f,
$$
we also have that:
$$
{\rm Var}[f(X)]
= \min_{c\in\mathbb{R}} \int \pi (f - c)^2.
$$

Note that the same story applies to $f\phi$ instead of just $f$ to deal with moments of order-1, order-2, and so on. But let's focus on~$f$ for simplicity.

As mentionned before, the VS asymptotic variance is the same as the IS variance, but when applied to $(f-g_\star)$ instead of $f$. Interestingly, the variable $f(X)-g_\star(X)$ has zero mean by the moment matching property inherited from KL~minimization.

Then comes a natural question: when is the variance of $(f-g_\star)$ smaller than that of $f$ under~$\pi$? Intuition suggests that, since $g_\star$ is the best fit to $f$ in the KL~sense, the residuals $f-g_\star$ tend to be ``small''. We can make this a little bit more precise using :
$$
\min_g \int \pi (f-g)^2 \leq \min_{c\in\mathbb{R}} \int \pi (f-c)^2 = {\rm Var}[f(X)].
$$

@@

The other option is to approximate the KL gradient altogether:
$$
\nabla D \approx {\cal I}[\pi(g_n - f)\phi].
$$

Option~2, which corresponds to our initial variational sampling idea, is arguably better because it is exact whenever $g_n=f$, meaning that the Newton iteration will converge deterministically whenever $f$ is in the approximating family. Hence, for a near Gaussian target, we expect less numerical instabilities with Option~2. Note that, at this level of description, there is no sampling kernel involved -- this is a specific detail of numerical integration.

What we have sketched here is nothing but a stochastic gradient method to minimize the KL~divergence, which is quite different in spirit from the initial VS~idea in that resampling could take place at each iteration of the optimization process. While this ensures that the successive gradient estimates are mutually independent, hence the method is unbiased and exact (not only asymptotically). This could be costly if the target factor takes time to evaluate, though.

\section{Probability updating}

What the variance inequalities essentially prescribe is that the target factor~$f$ be a bounded function. If this condition holds, the product distribution $\pi f$ does not have heavier tails than $\pi$, therefore IS converges whenever $\pi \phi\phi^\top$ exists, which is guaranteed for instance if $\pi$ is Gaussian (proper) and  $\phi$ is polynomial, which is our standard setting. VS convergence follows from the same argument provided that $(f-g)$ is bounded, in other words that the best fitting~$g$ is also bounded.

polynomial in the Gaussian case. In practice, this constraint is easily fulfilled, for instance if~$f$ is a posterior distribution arising from an exponential prior and a bounded likelihood function. Other compliant factors can be derived from this case, as we will see. However, importance weighting functions resulting from distribution ratios are ruled out.

Consider a Bayesian posterior distribution of the form $p=\pi L$, where $\pi$ is a prior and $L$ a likelihood function. It is natural to apply VS with $\pi$ as the sampling distribution and $f=p/\pi=L$ as the factor to be approximated. This means that we are trying to approximate the likelihood function by a Gaussian function in the context of the prior. Makes sense. This is the canonical approach exposed in my previous publications on VS.
